---
title: "tera"
vignette: >
  %\VignetteIndexEntry{tera}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

This vignette walks through some of the basics of rendering templates with Tera
in R. Users mainly interact with a Tera R6 object, which serves as a template 
library with encapsulated methods for rendering templates with a given context. 

A templating engine requires two things:

- a `template`, as you may have guessed, that includes variables and rendering 
  logic describing where and how to inject data, and
- a `context`, or a set of variables and values to be injected into the 
  template.

Templating syntax is described in the documentation for [Tera](https://keats.github.io/tera/).

```{r}
#| label: setup
library(tera)
```

## Usage

To get a feel for what `tera` can do, let's start with a simple "hello world"
example.

```{r}
#| label: hello-world
tera <- Tera$new()

tera$render_string(
  '<p>Hello {{ x }}. This is {{ y }}.</p>',
  x = "world",
  y = "tera"
)
```

The syntax and API should look pretty familiar to anyone who has used `glue` to
do something like `glue::glue("Foo { x }", x = "bar")`. The big difference is
the object-oriented workflow.

## Initializing `Tera`

Everything in `tera` revolves around the `Tera` object, which serves as a 
template library with encapsulated rendering methods. In the above example, we 
initialize a `Tera` engine with an empty template library by calling `Tera$new()` 
with no arguments.

If you have a complicated directory system with nested templates and inheritance 
patterns - a common situation for web development, you may find it easier to 
initialize an `tera` by specifying the directory with a glob pattern. Suppose, 
for example, that you have a website directory that looks like this: 

```{r}
#| label: directory
template_dir <- system.file("templates", package = "tera")

cat(
  "website",
  list.files(template_dir, recursive = TRUE),
  sep = "\n- "
)
```

You can generate a new `Tera` around this directory like so

```{r}
#| label: glob
tera <- Tera$new(file.path(template_dir, "**/*.html"))
tera
```

## Adding templates

Templates can be added from file or string.

```{r}
#| label: add-templates
# add templates manually from file
tera <- Tera$new()
tera$add_file_templates(
  "base.html" = file.path(template_dir, "base.html"),
  "index.html" = file.path(template_dir, "index.html")
)
tera

# add templates manually from string
tera$add_string_templates(
  img = '<img src="{{ img_src }}">'
)
tera
```

If you initialize a tera instance with a glob and then add more file templates 
to the template directory, you can reload with the glob to catch the new 
templates.

```{r}
#| label: reload
example_dir <- file.path(tempdir(), "tera-templates")
dir.create(example_dir)

file.copy(
  from = file.path(template_dir, "base.html"),
  to = file.path(example_dir, "base.html")
)

tera <- Tera$new(file.path(example_dir, "*.html"))
tera

file.copy(
  from = file.path(template_dir, "index.html"),
  to = file.path(example_dir, "index.html")
)

tera$reload()
```

Reset to continue with built-in template examples.

```{r}
#| label: reset-tera-instance
tera <- Tera$new(file.path(template_dir, "**/*.html"))
```

## Rendering basics

Generally speaking, rendering a template involves supplying it with a `context`, 
or a set of key-value pairs, with the keys being the variable names - surrounded 
by `{{{ variable }}}` in the template - and their values being the content to 
inject into the template. You can render a template to string or to a file. 
Consider our hypothetical website's index template:

```{r}
#| label: index-template
template_dir |>
  file.path("index.html") |>
  readLines(warn = FALSE) |>
  cat(sep = "\n")
```

This has four variables: `{{{ title }}}`, `{{{ x }}}`, `{{{ y }}}`, and 
`{{{ owner }}}`. You can render this template to a string by passing it a 
context, a set of values for those variables. Here we render the template to a 
string.

```{r}
#| label: render-to-string
string <- tera$render_template(
  "index.html",
  title = "This is my blog",
  p = "Welcome to my awesome homepage.",
  owner = "Blake"
)

cat(string)
```

You can also render a template to file by specifying `outfile`.

```{r}
#| label: render-to-file
outfile <- file.path(tempdir(), "rendered-index.html")

tera$render_template(
  "index.html",
  title = "This is my blog",
  p = "Welcome to my awesome homepage.",
  owner = "Blake",
  outfile = outfile
)

cat(readLines(outfile, warn = FALSE), sep = "\n")
```

And you can render a specific block in a template:

```{r}
#| label: render-block
string <- tera$render_template(
  "index.html",
  title = "This is my blog",
  p = "Welcome to my awesome homepage.",
  owner = "Blake",
  block = "content"
)

cat(string)
```

You can also render individual components defined in one of your templates, for
example

```{r}
#| label: render-component
string <- tera$render_component(
  "widget",
  title = "foo",
  body = "<p>This is a widget!</p>"
)

cat(string)
```

And you can bypass library templates altogether and pass a template string
directly

```{r}
#| label: render-string-template
string <- tera$render_string(
  '<img src="{{ img_src }}">',
  img_src = "foo/bar.svg"
)

cat(string)
```

Two helper functions are also provided if you want to render a one-off template
without going throught he process of initializing a tera instance. The 
`render_template()` function will render a template file, and `render_string()`
will render a string, same as the methods, but without the explicit tera 
instance.

## Inspecting the library

There are some tools for inspecting the library. You can get a list of 
templates and components in the library and a list of variables in a specific
template.

```{r}
#| label: inspection
tera$templates()
tera$components()
tera$variables("index.html")
```

There is also the print method, which provides some of this information.

```{r}
tera$print()
```

## Rendering logic

The `Tera` templating engine offers a lot of additional functionality, like 
control flow and data manipulation. For example, the blog post template shows 
how to construct a for loop and apply built-in filters and functions. 

```{r}
#| label: for-loop
template_dir |>
  file.path("blog", "post.html") |>
  readLines(warn = FALSE) |>
  cat(sep = "\n")
```

In `{{{ now() | date(format="%Y-%m-%d") }}}`, `now()` is a function that returns
the current date and time. It's returned value is then piped to the `date()` 
filter, which provides formatting options. In R, one way of achieving the same
thing is `format(Sys.Date(), "%Y-%m-%d") `. The template also has the for-loop 
construction `{$ for product in products %}` that allows for looping over the 
elements of a product table or array. When passed a data.frame, we get this:

```{r}
#| label: render-logic
products <- data.frame(
  name = c("apple", "banana", "orange"),
  price = c(0.25, 0.4, 0.5)
)

string <- tera$render_template(
  "blog/post.html",
  title = "Fruit prices",
  products = products
)

cat(string)
```

## Autoescape

Tera escapes HTML by default:

```{r}
#| label: autoescape
string <- tera$render_string(
  "{{ html }}",
  html = "<script>alert('Hello World!')</script>"
)

cat(string)
```

For one-off rendering like the above, the function takes an autoescape argument.

```{r}
#| label: autoescape-arg
string <- tera$render_string(
  "{{ html }}",
  html = "<script>alert('Hello World!')</script>",
  autoescape = FALSE
)

cat(string)
```

For rendering templates and components in the library, you can turn off 
autoescape globally using `$autoescape_off()` and turn it back on with 
`$autoescape_on()` (it is on by default).

```{r}
#| label: autoescape-toggle
tera$autoescape_off()
tera$autoescape_on()
```

## Delimiters

Templates are marked up with `{% %}` for blocks, `{{ }}` for variables, and
`{# #}` for comments. You can change these with `$set_delimiters()`, though only
on an engine with an empty template library, so it must be done before any
templates are added.

```{r}
#| label: set-delimiters
alt <- Tera$new()

alt$set_delimiters(
  variable_start = "<<",
  variable_end = ">>"
)

alt$render_string("<< greeting >>, world!", greeting = "Hello")
```

Each start delimiter must differ from the others, each delimiter must be exactly
two bytes long, and any delimiter left `NULL` is unchanged. Use `$delimiters()`
to see the current set.

```{r}
#| label: delimiters
alt$delimiters()
```

## Inheritance

Templates can inherit content from each other in one of two ways, either using
`include` or, for more complicated inheritance, `extends`.

```{r}
#| label: include
string <- tera$render_string(
  '{%- include "index.html" -%}',
  title = "This is my blog",
  p = "Welcome to my awesome homepage.",
  owner = "Blake"
)

cat(string)
```

The extension mechanism is a little more involved, requiring that you specify
content blocks where content from a child document should be injected. We have
actually been using this method in the examples already. Our current library has 
`base.html`:

```{r}
#| label: base.html
template_dir |>
  file.path("base.html") |>
  readLines(warn = FALSE) |>
  cat(sep = "\n")
```

Notice it has two `{% block ... %}`. This is where a child document inserts
content. And here is `index.html`:

```{r}
#| label: index.html
template_dir |>
  file.path("index.html") |>
  readLines(warn = FALSE) |>
  cat(sep = "\n")
```

Notice it has `{% extends "base.html" %}`. This makes it a child document of
`base.html`.

```{r}
#| label: extends
tera <- Tera$new(file.path(template_dir, "**/*.html"))

string <- tera$render_template(
  "index.html",
  title = "This is my blog",
  p = "Welcome to my awesome homepage.",
  owner = "Blake"
)

cat(string)
```
